An Intrinsically Typed Solution for the List Machine Benchmark

### 2020-06-23 papers/SBLP2020/paper.tex
More updates on related work

\paragraph{Benchmarks for PL mechanization}{
POPLMark chalenge~\cite{Aydemir05} .... Another chalenge was proposed by Pientka et. al.~\cite{Pientka18}...
}

\paragraph{Definitional intepreters}{
The use of definitional interpreters for specifying semantics dates back to Reynold's pioneer work~\cite{Reynolds72}.
Recently, the interest on such interpreters was revitalized by Amin et. al.~\cite{Amin17} which used definitional
interpreters, implemented in Coq, to prove type soundness theorems for non-trivial typed languages like System F$_{<:}$.
Unlike our work, Amin et. al. formalizations does not use intrinsically-typed representations of their syntax which cluttered
their formalization with the need of dealing with ``stuck states'' in the semantics. Poulsen el. al. described how
definitional interpreters for imperative languages can be much concisely implemented by using intrisincally-typed syntax by
using a library for name binding using \emph{scope graphs}~\cite{Poulsen18}, which greatly simplifies the treatment of complex
binding structures. Since, the list machine benchmark was designed to address other problems than binding, our representation
using \emph{de Bruijn} indexes was sufficient to implement the desired typechecker and interpreter.
Other recent application of definitional interpreters was proposed by Rouvoet et. al.~\cite{Rouvoet20}. The main contribution
of Rouvoet et. al. was to define interpreters for linear typed languages supporting session based concurrency. In order to model
linear typing features in the intepreter, authors have implemented monads based on a separation algebra which they
named \emph{Market}, supporting the main operations for accessing and update the store used by the intepreter.
Another work using intrisincally-typed syntax for resource control was developed by Thiemann~\cite{Thiemann19}, which
implemented an interpreter for a more realistic core functional session typed calculus including recursion and
session subtyping in Agda. Thiemann modeled the semantics as an interruptible abstract machine which provides a
simple interface to a scheduler.
}

@@

>>>>

\paragraph{Benchmarks for PL mechanization}{
The importante of benchmarks is recognized in several areas in computer science. A relevent example is the
TPTP library for the theorem proving community~\cite{Sut17}. In the context of programming languages community,
the POPLMark challenge~\cite{Aydemir05} was developed by a group renowned programming language researchers aiming
the collaboration between the PL community and the proof assistants researchers. The main objective of this challenge
was to motivate authors to formalize all of their theorems using such tools. Since the focus of the POPLMark challenge
was mainly the formalization of type soundness theorems, other benchmarks were proposed with different objectives. The list
machine benchmark was proposed by Appel et. al.~\cite{Appel07} as an exercise in formalizing results that interest
compiler oriented research and also provides Twelf and Coq solutions to this benchmark. Our work provides another
solution to this benchmark using intrinsically-typed syntax in Agda programming language. Representation of binding syntax
was the subject of Felty et. al.~\cite{FeltyMP18} which proposed a set of problems and research questions for tools
that use the high-order abstract syntax approach for name binding. Since the list machine benchmark definition avoids
name binding issues because they are orthogonal to most of compiler related proofs~\cite{Appel07}, we just ensure
the correct manipulation of names by following the traditional de Bruijn approach. Finally, recent problem set
was proposed by Pientka et. al.~\cite{Pientka18}, named POPLMark challenge reloaded, focusing on the mechanization
of logical relation arguments, like strong normalization theorems.}

\paragraph{Definitional intepreters}{
The use of definitional interpreters for specifying semantics dates back to Reynold's pioneer work~\cite{Reynolds72}.
Recently, the interest on such interpreters was revitalized by Amin et. al.~\cite{Amin17} which used definitional
interpreters, implemented in Coq, to prove type soundness theorems for non-trivial typed languages like System F$_{<:}$.
Unlike our work, Amin et. al. formalizations does not use intrinsically-typed representations of their syntax which cluttered
their formalization with the need of dealing with ``stuck states'' in the semantics. Poulsen el. al. described how
definitional interpreters for imperative languages can be much concisely implemented by using intrisincally-typed syntax by
using a library for name binding using \emph{scope graphs}~\cite{Poulsen18}, which greatly simplifies the treatment of complex
binding structures. Since, the list machine benchmark was designed to address other problems than binding, our representation
using \emph{de Bruijn} indexes was sufficient to implement the desired typechecker and interpreter.
Other recent application of definitional interpreters was proposed by Rouvoet et. al.~\cite{Rouvoet20}. The main contribution
of Rouvoet et. al. was to define interpreters for linear typed languages supporting session based concurrency. In order to model
linear typing features in the intepreter, authors have implemented monads based on a separation algebra which they
named \emph{Market}, supporting the main operations for accessing and update the store used by the intepreter.
Another work using intrisincally-typed syntax for resource control was developed by Thiemann~\cite{Thiemann19}, which
implemented an interpreter for a more realistic core functional session typed calculus including recursion and
session subtyping in Agda. Thiemann modeled the semantics as an interruptible abstract machine which provides a
simple interface to a scheduler. Since our objective was formalizing the list machine benchmark in Agda and it
does not have linearity constraints in its state manipulation, we do not need to deal with the complexities
of linear and session types as~\cite{Rouvoet20,Thiemann19}.}

@@

This paper shows the combination of intrinsically-typed syntax and definitional interpreters to formalize programming languages. Using such approach, we were able to provide a machine checked version of the list-machine benchmark in Agda, showing that the approach is useful to formalize both high-level and low-level languages. The ideas presented here can be exploited on the formalization of real-world virtual machines, such as the JVM and LUA VM, since we were able to encode features such as jumps, mutable state, and sub-typing. When comparing such approach with the conventional syntactic way to prove properties, we can affirm that it requires a small number of lines to achieve the same result, even without the use of proof automation. This happens because the approach uses the power of the host language, receiving some proofs for free.

As future work, we want to reuse the ideas presented in this paper to provide an intrinsically-typed formalization for languages like the eBPF and the LUA VM. Furthermore, one can extend the formalization presented here for other programming languages with similar settings.
